Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-15 manuscript/manuscript.tex
Edit the introduction.

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are simplified mathematical models, which connect historical quantities such as the population size to observable features of genetic data, and thus form the basis of parametric inference methods.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

One of the most important and widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to observable data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

The Kingman coalescent underlies a variety of methods for inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size, $N$~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Observable properties of genetic data depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains some features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or pattern of heterozygocity along the chromosome (e.g., \cite{LiDurbin2011}).

A serious problem for this class of inference procedures is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, when a beneficial mutation increases rapidly in frequency, it distorts the gene genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
When such selective sweeps are common, a large fraction of coalescence events may occur during sweeps.
In this case, the average coalescent rate is proportional to the beneficial mutation supply rate, which is itself \emph{directly} rather than inversely proportional to $N$.
Thus, the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of the neutral model underlying the Kingman coalescent may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.

In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be \emph{qualitatively} incorrect.

It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{SelectionScanReview}).
These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic history.
Genomic regions that are outliers from this distribution are presumed to be under selection.
In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.

@@

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
Finally, we demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

[STREAMLINE THE FOLLOWING:]

Consider a sample of $n$ chromosomes taken from a population of size $N$ with mutation rate $\mu$ per site per generation.
Following the notation of \cite{Fu1995}, the sample site frequency spectrum is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.

The SFS and folded SFS are single-site statistics.
As such, they can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.
We define the two-site frequency spectrum (2-SFS) at distance $d > 0$,
$\left\{ \xi_{ij}(d) : 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
By analogy to the folded SFS, we define the folded 2-SFS, as
\begin{align*}
    \eta_{ij}(d) &= \xi_{ij}(d) \\
                 &+ (1-\delta_{i,n-i}) \xi_{n-i,j}(d) \\
                 &+ (1-\delta_{j,n-j}) \xi_{i,n-j}(d) \\
                 &+ (1-\delta_{i,n-i}) (1-\delta_{j,n-j}) \xi_{n-i,n-j}(d).
\end{align*}
In this section, we will be considering non-recombining sites and so will suppress the distance $d$.

In the limit of low per-site mutation rate ($\mu\to0$), all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
    \label{eq:expected_2sfs}
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \neq i, (n-i)$.
This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \neq i, (n-i)$.
(There are positive correlations between mutations at complimentary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

@@

Our goal is to demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.
To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following useful transformation.

>>>>

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are simplified mathematical models, which connect unobserved quantities such as the population size to observable features of genetic data.
However, populations are complicated and, moreover, varied in their complications.
No simple model can capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982a, Kingman1982b, Kingman1982c, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies, trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model may be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982a}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982b}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygocity along a chromosome (e.g., \cite{LiDurbin2011}).

A serious problem for this class of inference methods is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
One of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For example, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly} rather than inversely proportional to the population size.
It follows that the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.

While the example above is extreme, it is well-established that violations of the neutrality assumption can distort or mask the signatures of population size changes.
For example, \cite{SchriderEtAl2016} recently demonstrated that several popular inference methods give misleading results in the presence of selective sweeps.
In a similar vein, \cite{CvijovicEtAl2018}, showed that purifying selection at linked sites that is sufficient to reduce genetic diversity is also sufficient to distort the SFS, leading to a false signal of population growth.
Moreover, genomic evidence from multiple species suggests that such violations of neutrality may be widespread \autocite{SellaEtAl2009, Corbett-DetigEtAl2015, KernHahn2018}.

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatscheck2013, DesaiEtAl},
and purifying selection at sufficiently dense sites \autocite{Seger, Good, Nicholaisen}.

In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.

It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{VittiEtAl2013}).
These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
Genomic regions that are outliers from this distribution are presumed to be under selection.
In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.

@@

Here, we propose that summary statistics based on the two-site frequency spectrum (2-SFS)---the generalization of the SFS to pairs of nearby sites \autocite{Hudson2001, FerrettiEtal2018})---are useful for distinguishing between the Kingman coalescent with population growth and multiple merger coalescents.
In particular, we demonstrate that a transformation of the 2-SFS that we refer to as \emph{frequency pointwise mutual information} (fpmi) is sensitive to multiple mergers but relatively invariant under time-varying coalescent rates.
We show that this is true for both perfectly linked sites and sites separated by a moderate genetic distance.
These statistics may be calculated efficiently from genomic single nucleotide polymorphism data.
Furthermore, they do not require phasing, recombination maps, or ancestral allele identification and are informative even with small sample sizes.
Together, these properties make the 2-SFS useful for demographic model-checking in a wide range of species.
We demonstrate this model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complimentary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.

In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
We define the 2-SFS,
$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011} in a model with selection.
We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies.

In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the branch length distribution by
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
    \label{eq:expected_2sfs}
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
(We suppress the distance, $d$, for non-recombining sites.)
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

\cite{Fu1995} calculated the first and second moments of the branch-length distribution for non-recombining infinite sites locus under the standard time-homogeneous Kingman coalescent.
He found that $\E{\tau_i \tau_j} < \E{\tau_i}\E{\tau_j}$ for all $j \not\in \{i, (n-i)\}$.
This result, combined with \eqs{eq:expected_sfs}{eq:expected_2sfs}, implies a negative correlation between mutations at different frequencies: trees generating a mutation with derived allele count $i$ are less likely than average to generate a second mutation with derived allele count $j \not\in \{i, (n-i)\}$.
(There are positive correlations between mutations at complementary frequencies induced by genealogies whose root node partitions the tree into subtrees of size $i$ and $n-i$.)

@@

In the following, we demonstrate that this distinction in the 2-SFS between the multiple merger and Kingman coalescents
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.

% To this end, we will make use of the fact that the 2-SFS may be interpreted as a joint probability distribution to define the following transformation.
